Association between trauma severity models and opportunities for improvement:

### 2022-12-20 manuscript.Rmd
missing data

---
title: 'Association between trauma severity models and opportunities for improvement:
  A retrospective cohort study'
author: "Victoria Witthoff"
output:
  bookdown::pdf_document2: default
  bookdown::html_document2: default
  bookdown::word_document2: default
bibliography: bibliography.bib
csl: vancouver.csl
fontsize: 11pts
linestretch: 1
always_allow_html: yes
editor_options: 
   chunk_output_type: inline
---

@@

Participants
------------
A total of `r exclusion[1,2]` patients were recorded in the trauma registry during the study period. Of these, `r exclusion[2,2]` were reviewed and therefore became eligible for study inclusion. `r exclusion[7,3]` patients were excluded, due to either being of paediatric age (`r exclusion[3,3]`), being dead on arrival at hospital (`r exclusion[4,3]`), or due to incomplete data (`r exclusion[5,3]`). Figure \@ref(fig:exclusion) shows the inclusion process for eligible participants.

```{r exclusion, fig.cap = "Flow diagram for application of exclusion criteria", fig.align = "center", echo=FALSE, out.width = "600px"}
knitr::include_graphics("images/exclusion.png") 
## exclusion_flowchart.R; save_png(exclusion_flowchart(exclusion), "exclusion") to update
```

Detailed totals of missing data for each required parameter in cases that were excluded for incomplete data are given in table \@ref(tab:incomplete-table). Some patients had missing data in several parameters and therefore the total number of missing parameters is greater than the total number of patients excluded due to incomplete data.

```{r incomplete-table, echo=FALSE, fig.align = "center"}
knitr::kable(table(na.data, type = "incomplete"), caption = "Parameters that were missing in cases excluded for incomplete data") %>% kable_styling(position = "center", full_width = FALSE)
```

@@

```{r table-one} 
  table.one
## this thing breaks the table counting but ill just fix it manually in word
```

@@

Key results
-----------
<!-- Summarise key results with reference to study objectives -->

This aim of this study was to evaluate whether several commonly used trauma severity models (TRISS, NORMIT, and PS) are associated with, and to what extent they can predict, OFI, with the goal of improving case selection for multidisciplinary morbidity and mortality conferences. The three models investigated did not differ significantly in their results. The main findings indicate that increased probability of survival was associated with lower odds for OFI, with odds ratios ranging from `r t['or']` to `r n['or']`. The AUC values were almost identical for all three models (range: `r p[['auc']]`-`r t[['auc']]`), indicating poor diagnostic ability in discriminating between OFI and no OFI. Calibration was also low for all models (ICI range: `r n[['ici']]` - `r p[['ici']]`). Accuracy at a cutoff of 0.5 was low across the board (range: `r p[['acc']]`-`r n[['acc']]`).

@@

There were several limitations to this study. One of them is that selection of cases for review at conferences was in large part based upon the use of audit filters. There is as yet no widespread consensus on which audit filters are the most effective at identifying cases and this is an ongoing topic of research. The choice of audit filters used affects case selection and in turn introduces an element of selection bias into this study. It is possible, and perhaps even likely, that some cases with opportunities for improvement were never selected for case review and therefore weren't available for inclusion in this study. Another limitation is that, due to the long duration of enrolment in this study, the composition of the staff present at conferences will have differed over time. Determining whether OFI exists or not is an essentially subjective evaluation and the conclusions reached by any given conference may vary depending on which individuals are present. It is also possible that standards and expectations of care have changed over time as knowledge and technology improves, which would also in turn affect assignment of OFI. <!--JA: I takt med att man åtgärdar systematiskta fel så upkommer na -->

@@

### Health equity
<!-- Is this section requested by KI?-->
The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide. `r dd['male']`% of participants were male and the median age was `r dd['age']`, which matches with statistics both in Sweden and globally [@globalburden]. The predominance of male patients invites one to question if female patients are less likely to receive quality care. In healthcare as in all areas, repetition leads to mastery of a skill, so an onlooker might be tempted to assume that the high volume of male patients in trauma will lead healthcare providers to superiority iin treating males, whilst treatment of female patients will be neglected. This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes. None of these studies found a correlation beween sex and increased mortality or other serious outcomes [@croce2002; @magnotti2008; @pape2019; @trentzsch2014], and in fact one study in the Netherlands found that female sex was correlated with decreased mortality [@pape2019].

>>>>

---
title: 'Association between trauma severity models and opportunities for improvement:
  A retrospective cohort study'
author: "Victoria Witthoff"
output:
  bookdown::word_document2: default
  bookdown::html_document2: default
  bookdown::pdf_document2: default
bibliography: bibliography.bib
csl: vancouver.csl
fontsize: 11pts
linestretch: 1
always_allow_html: yes
editor_options: 
   chunk_output_type: inline
---

@@

Participants
------------
A total of `r exclusion[1,2]` patients were recorded in the trauma registry during the study period. Of these, `r exclusion[2,2]` were reviewed and therefore became eligible for study inclusion. `r exclusion[7,3]` patients were excluded, due to either being of paediatric age (`r exclusion[3,3]`), being dead on arrival at hospital (`r exclusion[4,3]`), or due to incomplete data (`r exclusion[5,3]`). Figure \@ref(fig:exclusion) shows the inclusion process for eligible participants. Some patients had missing data in several parameters and therefore the total number of missing parameters is greater than the total number of patients excluded due to incomplete data.

```{r exclusion, fig.cap = "Flow diagram for application of exclusion criteria", fig.align = "center", echo=FALSE, out.width = "600px"}
knitr::include_graphics("images/exclusion.png") 
## exclusion_flowchart.R; save_png(exclusion_flowchart(exclusion, na.data), "exclusion") to update
```

@@

```{r table-one, centering = TRUE}
t1kable(table.one)
##knitr::kable(t1kable(table.one), caption = "Comparison of model results") %>% ##kable_styling(position = "center", full_width = FALSE)
```

@@

Key results
-----------
This aim of this study was to evaluate whether several commonly used trauma severity models (TRISS, NORMIT, and PS) are associated with, and to what extent they can predict, OFI, with the goal of improving case selection for multidisciplinary morbidity and mortality conferences. The three models investigated did not differ significantly in their results. The main findings indicate that increased probability of survival was associated with lower odds for OFI, with odds ratios ranging from `r t['or']` to `r n['or']`. The AUC values were almost identical for all three models (range: `r p[['auc']]`-`r t[['auc']]`), indicating poor diagnostic ability in discriminating between OFI and no OFI. Calibration was also low for all models (ICI range: `r n[['ici']]` - `r p[['ici']]`). Accuracy at a cutoff of 0.5 was low across the board (range: `r p[['acc']]`-`r n[['acc']]`).

@@

There were several limitations to this study. One of them is that selection of cases for review at conferences was in large part based upon the use of audit filters. There is as yet no widespread consensus on which audit filters are the most effective at identifying cases and this is an ongoing topic of research. The choice of audit filters used affects case selection and in turn introduces an element of selection bias into this study. It is possible, and perhaps even likely, that some cases with opportunities for improvement were never selected for case review and therefore weren't available for inclusion in this study. Another limitation is that, due to the long duration of enrolment in this study, the composition of the staff present at conferences will have differed over time. Determining whether OFI exists or not is an essentially subjective evaluation and the conclusions reached by any given conference may vary depending on which individuals are present. It is also possible that standards and expectations of care have changed over time as knowledge and technology improves, which would also in turn affect assignment of OFI.

@@

### Health equity
The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide. `r dd['male']`% of participants were male and the median age was `r dd['age']`, which matches with statistics both in Sweden and globally [@globalburden]. The predominance of male patients invites one to question if female patients are less likely to receive quality care. In healthcare as in all areas, repetition leads to mastery of a skill, so an onlooker might be tempted to assume that the high volume of male patients in trauma will lead healthcare providers to superiority iin treating males, whilst treatment of female patients will be neglected. This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes. None of these studies found a correlation beween sex and increased mortality or other serious outcomes [@croce2002; @magnotti2008; @pape2019; @trentzsch2014], and in fact one study in the Netherlands found that female sex was correlated with decreased mortality [@pape2019].

### 2022-12-20 pub.Rmd
missing data

---
title: "Association between trauma severity models and opportunities for improvement:
  A retrospective cohort study"
output: html_document
author: "Victoria Witthoff"
bibliography: bibliography.bib
csl: vancouver.csl
---

@@

```{r main}
  ## source("main.R") 
```

@@

You can include code in this document like this:

```{r main, echo=FALSE}
## source("main.R") ## This "imports" the main script file of your project and run any code in it
```

You can also embed plots:

```{r plot, echo=FALSE}
plot(pressure)
```

>>>>

---
title: 'Association between trauma severity models and opportunities for improvement:
  A retrospective cohort study'
author: "Victoria Witthoff"
output:
  bookdown::pdf_document2: default
  bookdown::html_document2: default
  bookdown::word_document2: default
bibliography: bibliography.bib
csl: vancouver.csl
fontsize: 11pts
linestretch: 1
always_allow_html: yes
editor_options: 
   chunk_output_type: inline
---

@@

```{r main}
  ## source("main.R") 
```
<!-- rmarkdown::render("manuscript.Rmd", html_document2(global_numbering = TRUE)) -->

@@
